---
title: "Indian House Price Prediction Project"
author:
  - name: "Mentor: Urbi Datta"
  - name: "Group members: Atreya Ghoshal, Bighnesh Sahu, Srijita Datta, Sumit Jha,            Swarnava Kundu"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false    
editor: visual
---

## File description:

The **"India Housing Prices"** dataset contains 2.5 lakhs rows and 23 columns, offering detailed insights into housing market trends across various Indian states. It includes attributes related to - property types, pricing, location, and amenities, making it suitable for analysis ranging from basic descriptive statistics to advanced machine learning applications.

```{r}
dt = read.csv(file = "D:/MTTS STDSMLAI/Indian_House_Pricing_Project/india_housing_prices.csv", 
              header = TRUE)
#head(dt,n=3)
```

```{r}
dim(dt)
```

```{r}
colnames(dt)
```

## Columns in our dataset:

Some preliminary observations of the given dataset:

-   **ID**: Different rows having different Integer values.

-   **State**: Indian States ( character ).

-   **City**: Name of Cities ( character ).

-   **Locality**: Name of Localities ( character ).

-   **Property_Type**: There are 3 types of properties - Apartment, Independent House, Villa ( character ).

-   **BHK**: 5 varieties are there - 1,2,3,4,5 ( integer ).

-   **Size_in_sqFt**: Plot size in Sqft ( integer ).

-   **Price_in_Lakhs**: Price of the properties in Lakhs ( Double ).

-   **Price_per_SqFt**: Price of the properties per Sqft ( Double ).

-   **Year_Built**: Year in which the property was built ( integer ).

-   **Furnished_Status**: Status of furnishing - Furnished, Unfurnished, Semi-furnished ( character ).

-   **Floor_No.**: Floor No. ( Integer ).

-   **Total_Floors**: Total no. of floors ( Integer ).

-   **Age_of_Property**: Age of the property ( Integer ).

-   **Nearby_Schools**: No. of schools nearby ( Integer ).

-   **Nearby_Hospitals**: No. of hospitals nearby ( Integer ).

-   **Public_Transport_Accessibility**: Ease of accessing public transportation - High, Low, Medium ( character ).

-   **Parking_Space**: Availability of Parking space - Yes, No ( character ).

-   **Security**: Proper security system - Yes, No ( character ).

-   **Amenities**: Amenities like: Playground, Gym, Pool, etc. ( character ).

-   **Facing**: Orientation of the house - East, West, South, North ( character ).

-   **Owner_Type**: Type of owner - Owner, Builder, Broker ( character ).

-   **Availability_Status**: Status of availability - Ready_to_move, Under_construction ( character ).

```{r}
typeof(dt$Availability_Status)
length(unique(dt$Availability_Status))
```

```{r}
summary(dt$Size_in_SqFt)
summary(dt$Price_in_Lakhs)
```

-   Size in square foot varies from 500 ft. to 5000 ft. with mean 2750 ft.

-   Price in lakhs varies from 10 lakhs to 500 lakhs with mean 254.6 lakhs.

```{r}
hist(dt$Size_in_SqFt, breaks = 30,probability = TRUE,
       main="Distribution of properties according to sizes", 
     xlab="Size (in sq ft)", ylim= c(0,0.00030))


```

```{r}
summary(dt$Price_in_Lakhs)
val = dt$Price_in_Lakhs/10
summary(val)
boxplot(dt$Price_in_Lakhs,horizontal = TRUE, 
        main = "Distribution of properties according to prices",
        xlab = "Price (Lakhs)")
hist(dt$Price_in_Lakhs, breaks = 20,probability = TRUE,
       main="Distribution of properties according to prices", 
     xlab="Price (in Lakhs)")
```

```{r}
summary(dt$Price_per_SqFt)
boxplot(dt$Price_per_SqFt,horizontal = TRUE,
        main = "Distribution of Property prices per sqft.",
        xlab = "Price (per sqft)")
hist( dt$Price_per_SqFt, breaks = 30,
    main="Distribution of properties w.r.t. prices per sqft.", 
    xlab="Price (per sqft.)", ylim= c(0,80000))
```

-   The data is **positively skewed (right-skewed).**

-   The majority of the properties are clustered at the lower end of the price scale (between 0.0 and 0.2).

-   Here, a long tail is extending towards the right. This indicates that while most properties are relatively affordable (according to this scale), there are a few properties with significantly higher prices per square foot.

## Observations from the data

### Pricing

-   **Price_in_Lakhs = Size_in_SqFt × Price_per_SqFt** + (Some dynamic charge/ discount )

```{r}
Price_diff <- dt$Price_in_Lakhs - (dt$Size_in_SqFt * dt$Price_per_SqFt)
typeof(dt$Size_in_SqFt)
pos_price_diff = sum(round(Price_diff,4) > 0.0)
no_price_diff = sum(round(Price_diff,2) == 0.0)
neg_price_diff = sum(round(Price_diff,4) < 0.0)
print((pos_price_diff+no_price_diff+neg_price_diff))
```

50.29% of the data has prices greater than the equation.

49.65% of the data has prices less than the equation.

0.05% of the data has exactly same value as the equation.

```{r}
# Predicting the rate (Price per SqFt) instead of total price
rate_model <- lm(Price_per_SqFt ~ Size_in_SqFt, data = dt)

# Plotting the relationship
plot(dt$Size_in_SqFt, dt$Price_per_SqFt, 
     pch = 16, col = rgb(0,0,1,0.1),
     main = "Relationship: Size vs Price per sqft",
     xlab = "Size in sqft", ylab = "Price per sqft")
abline(rate_model, col = "red", lwd = 2)
```

-   The plot displays a **strong negative non-linear relationship**, where the price per square foot drops sharply as property size increases before stabilizing for larger homes.

### Locality

```{r}
#| warning: false
library(dplyr)
locality_summary <- dt %>%
  group_by(City) %>%
  summarize(
    localities = paste(unique(Locality), collapse = ", "),
    no_localities = length(unique(Locality))
  )
print(head(locality_summary,5))

```

After grouping by state and city, by checking randomly, we see that Locality_463 is present in Ahmedabad and Bhopal.

Now we do a quick search to find in which cities it is present.

```{r}
dt %>%
  filter(Locality == "Locality_463") %>%
  distinct(City)
```

-   We find that this locality is present in all of the 42 cities.

-   **Each city has 500 localities .**

```{r}
length(unique(dt$Locality))
```

-   There are 500 localities in total and we observe that all 500 localities are common in all the cities.

### Year Built

```{r}
length(unique(sort(dt$Year_Built)))
summary(dt$Year_Built)
```

-   We have properties built between 1990 to 2023.

```{r}
# Count houses per year
yearly_counts <- table(dt$Year_Built)
years <- as.numeric(names(yearly_counts))
counts <- as.vector(yearly_counts)


plot(years, counts, 
     type = "o", col = "darkred",pch = 16, lwd = 2,
     main = "Houses Built per Year",xlab = "Year",ylab = "Number of Houses",
     ylim = c(7150,7600))    
```

### BHK

This section examines the distribution of BHK configurations to identify the most frequently occurring housing type in the dataset.

```{r}
bhk_dist <- table(dt$BHK)

par(mar = c(5, 6, 4, 2),mgp = c(4, 1, 0))
bp <- barplot(bhk_dist, main = "Distribution w.r.t. BHK", xlab = "BHK Type", 
              ylab = "Number of Properties", 
              col = "darkorange",ylim = c(48000, 51000),xpd = FALSE, border = "black",las = 1)                


text(x = bp,y = bhk_dist,label = bhk_dist,pos = 3, cex = 0.9, font = 2,xpd = TRUE)      
```

```{r}
most_common_bhk <- names(which.max(bhk_dist))
most_common_bhk
```

-   The results indicate that the most common BHK configuration in the dataset is **1 BHK**.

### Property

```{r}
#Counting of properties each cities have
Property_list = dt %>%
  group_by(City) %>%
  summarise(count_prop = length(unique(ID)))
head(Property_list,5)
```

```{r}
Property_list$City[which.max(Property_list$count_prop)]
```

-   So, **Coimbatore** has maximum properties.

```{r}
# Property type wise average price
Price_per_prop= aggregate(Price_in_Lakhs ~ Property_Type + Availability_Status, data = dt, FUN = mean)
Price_per_prop
```

```{r}
# Property type wise average price per sqft
Price_per_sqft_prop= aggregate(Price_per_SqFt ~ Property_Type + Availability_Status, data = dt, FUN = median)
Price_per_sqft_prop
```

```{r}
property_type_dist <- table(dt$Property_Type)


par(mar = c(5, 6, 4, 2),mgp = c(4, 1, 0))

bp = barplot(property_type_dist,
        main = "Distribution of Property Types",
        xlab = "Property Type",
        col = "steelblue",xpd = FALSE, ylim = c(80000,85000),
        las = 1)

text(x = bp, 
     y = property_type_dist, 
     labels = property_type_dist, 
     pos = 3, xpd = FALSE,
     cex = 0.8, 
     font = 2)
```

### Impact of Public Transport Accessibility on Property Prices

```{r}
pt_price <- aggregate(
  Price_in_Lakhs ~ Public_Transport_Accessibility,
  data = dt,
  median
)

pt_price
```

-   The accessibility of public transport has minimal effect on prices.

### Furnished vs Unfurnished Property Prices

```{r}
furnish_price <- aggregate(
  Price_in_Lakhs ~ Furnished_Status,
  data = dt,median
)

furnish_price
```

-   According to the data the furnishing status has minimal effect on prices.

### Year Wise Price Change Analysis

```{r}
data_1990 = dt[dt$Year_Built == 1990,]
summary(data_1990$Price_in_Lakhs)
hist(data_1990$Price_in_Lakhs,breaks = 30,probability = TRUE,
     main="Distribution of prices of properties built in 1990", 
    xlab="Price (in lakhs)", ylim= c(0,0.0030))
```

```{r}
data_1991 = dt[dt$Year_Built == 1991,]
summary(data_1991$Price_in_Lakhs)
hist(data_1991$Price_in_Lakhs,breaks = 30,probability = TRUE,
     main="Distribution of prices of properties built in 1991", 
    xlab="Price (in lakhs)" , ylim= c(0,0.0030))

```

-   The summary of prices for these two years are similar.

```{r}
# Calculate the yearly statistics
yearly_mean <- aggregate(Price_in_Lakhs ~ Year_Built, data = dt, FUN = mean)
yearly_min  <- aggregate(Price_in_Lakhs ~ Year_Built, data = dt, FUN = min)
yearly_max  <- aggregate(Price_in_Lakhs ~ Year_Built, data = dt, FUN = max)
yearly_median <- aggregate(Price_in_Lakhs ~ Year_Built, data = dt, FUN = median)
```

-   

    ```{r}
    par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
    # initial plot with the Max (red)
    plot(x = yearly_max$Year_Built, yearly_max$Price_in_Lakhs, 
          type = "o", col = "red", lty = 2, pch = 18,
          xlab = "Year Built", 
         ylab = "Price (in Lakhs)",
         main = "Yearly Price Trends: Max")

    # mean line
    plot(yearly_mean$Year_Built, yearly_mean$Price_in_Lakhs, 
         type = "o", 
         pch = 19,
         col = "darkgreen",
         xlab = "Year Built", 
         ylab = "Price (in Lakhs)",
         main = "Yearly Price Trends: Mean")

    # Minimum line (Dashed blue)
    plot(yearly_min$Year_Built, yearly_min$Price_in_Lakhs, 
          type = "o", col = "blue", lty = 2, pch = 18,
          xlab = "Year Built", 
         ylab = "Price (in Lakhs)",
         main = "Yearly Price Trends: Min")
    ```

-   The mean price fluctuates within a very narrow band (roughly between **251 and 257 Lakhs**). This tells us that despite individual year-to-year variations, the house prices didn't change drastically over the years.

```{r}
#| warning: false
#For Ready_to_Move House
yearly_mean_ready <- aggregate(Price_in_Lakhs ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Ready_to_Move"), ], FUN = mean)

#For Under_Construction House
yearly_mean_cons <- aggregate(Price_in_Lakhs ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Under_Construction"), ], FUN = mean)

library(plotly)

p <- plot_ly() %>%
  # Line for Ready to Move
  add_trace(data = yearly_mean_ready, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Ready_to_Move',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'darkgreen')) %>%
  
  # Line for Under Construction
  add_trace(data = yearly_mean_cons, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Under_Construction',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'blue', dash = 'dash')) %>%
  
  # Line for Overall Median
  add_trace(data = yearly_mean, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Overall Mean',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'red', dash = 'dash')) %>%
  
  # Layout and Axis Titles
  layout(title = "Built-Year wise Mean Price Trends: Ready_to_Move vs Under_Construction",
         xaxis = list(title = "Year Built"),
         yaxis = list(title = "Price (in lakhs)", range = c(240, 270)),
         hovermode = "x unified") # Shows all values when hover over a year

# Display the plot
p


```

```{r}
### Calculation for median
yearly_median_ready <- aggregate(Price_in_Lakhs ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Ready_to_Move"), ], FUN = median)

#For Under_Construction House
yearly_median_cons <- aggregate(Price_in_Lakhs ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Under_Construction"), ], FUN = median)


p1 <- plot_ly() %>%
  # Line for Ready to Move
  add_trace(data = yearly_median_ready, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Ready_to_Move',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'darkgreen')) %>%
  
  # Line for Under Construction
  add_trace(data = yearly_median_cons, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Under_Construction',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'blue', dash = 'dash')) %>%
  
  # Line for Overall Median
  add_trace(data = yearly_median, x = ~Year_Built, y = ~Price_in_Lakhs, name = 'Overall Median',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'red', dash = 'dash')) %>%
  
  # Layout and Axis Titles
  layout(title = "Built-Year wise Median Price Trends:<br>Ready_to_Move vs Under_Construction",
         xaxis = list(title = "Year Built"),
         yaxis = list(title = "Price (in Lakhs)", range = c(240, 270)),
         hovermode = "x unified") # Shows all values when hover over a year

# Display the plot
p1


```

```{r}
### Calculation for Median
yearly_median_sqft_ready <- aggregate(Price_per_SqFt ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Ready_to_Move"), ], FUN = median)

#For Under_Construction House
yearly_median_sqft_cons <- aggregate(Price_per_SqFt ~ Year_Built,data = 
                                 dt[which(dt$Availability_Status  =="Under_Construction"), ], FUN = median)

#For Median Price line
yearly_median_sqft <- aggregate(Price_per_SqFt ~ Year_Built, data = dt, FUN = median)


p2 <- plot_ly() %>%
  # Line for Ready to Move
  add_trace(data = yearly_median_sqft_ready, x = ~Year_Built, y = ~Price_per_SqFt, name = 'Ready_to_Move',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'darkgreen')) %>%
  
  # Line for Under Construction
  add_trace(data = yearly_median_sqft_cons, x = ~Year_Built, y = ~Price_per_SqFt, name = 'Under_Construction',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'blue', dash = 'dash')) %>%
  
  # Line for Overall Median
  add_trace(data = yearly_median_sqft, x = ~Year_Built, y = ~Price_per_SqFt, name = 'Overall Median',
            type = 'scatter', mode = 'lines+markers', line = list(color = 'magenta', dash = 'dash')) %>%
  
  # Layout and Axis Titles
  layout(title = "Built-Year wise Median Price per SqFt Trends: Ready_to_Move vs Under_Construction",
         xaxis = list(title = "Year Built"),
         yaxis = list(title = "Price (per SqFt)", range = c(0.085,0.105)),
         hovermode = "x unified") # Shows all values when hover over a year

# Display the plot
p2
```

### Orientation of the house

```{r}
#Counting of houses facing each Direction
facing_counts <- table(dt$Facing)
par(mar = c(5, 6, 4, 2),mgp = c(4, 1, 0))

bp1 = barplot(table(dt$Facing), 
        main = "Property distribution according to orientation",
        col = "steelblue",
        xlab = "Direction",
        ylab = "Number of houses",  ylim= c(60000,64000),xpd = FALSE, 
        las = 1) 

text(x = bp1, 
     y = facing_counts, 
     labels = facing_counts, 
     pos = 3, xpd = FALSE,
     cex = 0.8, 
     font = 2)
```

```{r}
#The average prices of houses based on their facing direction
avg_price_summary <- dt %>%
  group_by(Facing) %>%
  summarise(Average_Price = mean(Price_in_Lakhs),
            Property_Count = n()) %>%
  arrange(desc(Average_Price))

print(avg_price_summary)
```

```{r}
# Generate frequency tables for all four directions
n_counts <- table(dt$Year_Built[dt$Facing == "North"])
s_counts <- table(dt$Year_Built[dt$Facing == "South"])
e_counts <- table(dt$Year_Built[dt$Facing == "East"])
w_counts <- table(dt$Year_Built[dt$Facing == "West"])


max_val <- max(c(n_counts, s_counts, e_counts, w_counts))


library(plotly)

plot_data <- dt %>%
  group_by(Year_Built, Facing) %>%
  summarise(Count = n(), .groups = 'drop')

# Creating the interactive plot
p <- plot_ly(plot_data, 
             x = ~Year_Built, 
             y = ~Count, 
             color = ~Facing, 
             colors = c("blue", "red", "darkgreen", "orange"),
             type = 'scatter', 
             mode = 'lines+markers') %>%
  layout(title = "Year-wise property sales acc. to orientation",
         xaxis = list(title = "Year Built"),
         yaxis = list(title = "Number of houses sold", range = c(1700, max(plot_data$Count) + 100)),
         legend = list(title = list(text = 'Click to Toggle Direction')))

# Display the plot
p
```

### Age of Property

```{r}
# Calculate the age statistics
age_mean <- aggregate(Price_in_Lakhs ~ Age_of_Property, data = dt, FUN = mean)
age_min  <- aggregate(Price_in_Lakhs ~ Age_of_Property, data = dt, FUN = min)
age_max  <- aggregate(Price_in_Lakhs ~ Age_of_Property, data = dt, FUN = max)
age_median <- aggregate(Price_in_Lakhs ~ Age_of_Property, data = dt, FUN = median)

plot(age_mean$Age_of_Property, age_mean$Price_in_Lakhs, 
     type = "o", lty = 1,
     pch = 19,
     col = "darkgreen",
     xlab = "Age of property", 
     ylab = "Price (in lakhs)",ylim=c(246,259),
     main = "Age of property price trends: Mean vs Median")

lines(age_median$Age_of_Property, age_median$Price_in_Lakhs, 
     type = "o", lty = 2,
     pch = 19,
     col = "blue")

#Legend
legend("bottomright", 
       legend = c("Mean", "Median"),
       col = c("darkgreen", "blue"), 
       lty = c(1, 2), 
       pch = c(19, 19,19),
       bty = "n")
```

-   There is no consistent trend showing that prices increase or decrease with the age of the property, suggesting that age alone does not explain the variation of price and more analysis is required.

### Amenities

```{r}
#| warning: false
# install.packages("fastDummies")
library(fastDummies)

# split by ", " to create individual columns
dt_expanded <- dummy_cols(dt, 
                          select_columns = "Amenities", 
                          split = ", ")

dt_expanded$Amenities <- NULL

```

Extended dataset created.

```{r}
# Identify our amenity columns
amenity_cols <- c("Amenities_Garden", "Amenities_Clubhouse",
                  "Amenities_Pool", "Amenities_Gym", "Amenities_Playground","Parking_Space","Security") 

# Map "Yes" to 1 and "No" to 0
dt_expanded$Security <- ifelse(dt_expanded$Security == "Yes", 1, 0)
dt_expanded$Parking_Space <- ifelse(dt_expanded$Parking_Space == "Yes", 1, 0)

# Sum the columns to get the count of houses for each
amenity_counts <- colSums(dt_expanded[amenity_cols], na.rm = TRUE)

# summary table
amenity_summary <- data.frame(
  Amenity = names(amenity_counts),
  Count = as.numeric(amenity_counts),
  Percentage = round((as.numeric(amenity_counts) / nrow(dt_expanded)) * 100, 2)
)
print(amenity_summary)
```

```{r}
# Amenity columns
cols_to_compare <- c("Price_in_Lakhs", "Price_per_SqFt",
                     "Amenities_Garden", "Amenities_Clubhouse",
                  "Amenities_Pool", "Amenities_Gym", "Amenities_Playground")

#  Calculate the matrix
cor_matrix <- cor(dt_expanded[cols_to_compare])

# Correlation Plot
corrplot::corrplot(cor_matrix, 
         method = "color",       
         addCoef.col = "black", 
         number.cex = 0.7,       
         type = "upper",         
         diag = FALSE)

```

-   Correlation between amenities and prices is 0 in most of the cases.

```{r}
#Price for houses with amenity pool versus those without it
amenity_stats_pool <- aggregate(Price_in_Lakhs ~ Amenities_Pool, data = dt_expanded, 
                           FUN = function(x) c(Mean = mean(x), 
                                               Median = median(x), 
                                               SD = sd(x), 
                                               Min = min(x), 
                                               Max = max(x)))

# Convert to a readable data frame
amenity_stats_pool <- do.call(data.frame, amenity_stats_pool)
print(amenity_stats_pool)
```

-   Presence of pool has little impact on our data.

```{r}
# Visualizing the impact of having a Gym on Price
boxplot(Price_in_Lakhs ~ Amenities_Gym , data = dt_expanded,
        main = "Price Distribution: Availability of Gym",
        xlab = "Gym presence (1 = Yes, 0 = No)",
        ylab = "Price (in lakhs)",
        col = c("lightgrey", "gold"))
```

-   We observe that the availability of gym also has little impact on the prices and so do the other amenities.

```{r}

calculate_premium <- function(amenity_name) {
  # Mean price with the amenity
  mean_with <- mean(dt_expanded$Price_in_Lakhs[dt_expanded[[amenity_name]] == 1], na.rm = TRUE)
  
  # Mean price without the amenity
  mean_without <- mean(dt_expanded$Price_in_Lakhs[dt_expanded[[amenity_name]] == 0], na.rm = TRUE)
  
  # Calculate Premium
  premium_val <- mean_with - mean_without
  premium_pct <- (premium_val / mean_without) * 100
  
  
  data.frame(
    Amenity = amenity_name,
    Avg_Price_With = round(mean_with, 2),
    Avg_Price_Without = round(mean_without, 2),
    Premium_Lakhs = round(premium_val, 2),
    Premium_Percent = round(premium_pct, 2)
  )
}


premium_table <- do.call(rbind, lapply(amenity_cols, calculate_premium))


premium_table <- premium_table[order(-premium_table$Premium_Percent), ]

print(premium_table)
```

-   We can observe that the presence of amenities affect the prices very little. But Playground has negative premium.

### City

```{r}
#State wise city count
State_summary <- dt %>% 
  group_by(State) %>%  
      summarize(Cities = paste(unique(City), collapse = ", "),
                no_cities = length(unique(City)))
```

```{r}
# Calculate average price per city
city_summary <- dt %>%
  group_by(City) %>%
  summarise(
    Avg_Price_Lakhs = mean(Price_in_Lakhs, na.rm = TRUE),
    Property_Count = n()
  )

print(city_summary)

library(tidygeocoder)

# For finding the Lat/Long for each city automatically
city_geo <- city_summary %>%
  geocode(city = City, method = 'osm', lat = latitude , long = longitude)

#BY default Geocode treats Vishakhapatnam as Vizag and Dwarka as
# the ancient city  of Gujrat
# Manually update Visakhapatnam and Dwarka coordinates in city_geo dataframe
city_geo <- city_geo %>%
  mutate(
    latitude = case_when(
      City == "Dwarka" ~ 28.5823,
      City == "Vishakhapatnam" ~ 17.6868,
      TRUE ~ latitude
    ),
    longitude = case_when(
      City == "Dwarka" ~ 77.0500,
      City == "Vishakhapatnam" ~ 83.2185,
      TRUE ~ longitude
    )
  )

#Creating interactive Map
library(leaflet)

pal <- colorNumeric(
  palette = "RdYlGn",
  domain = city_geo$Avg_Price_Lakhs,
  reverse = TRUE
)

leaflet(city_geo) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude, 
    lat = ~latitude,
    radius = 5,
    weight = 1,
    fillOpacity = 0.8,
    color = "white",
    fillColor = ~pal(Avg_Price_Lakhs),
    popup = ~paste0("<b>City:</b> ", City, 
                    "<br><b>Avg Price:</b> ", round(Avg_Price_Lakhs, 2), " Lakhs",
                    "<br><b>Total Properties:</b> ", Property_Count),
    clusterOptions = markerClusterOptions()
  ) %>%
  # Add a legend to make the colors clear
  addLegend(
    pal = pal, 
    values = ~Avg_Price_Lakhs, 
    opacity = 0.7, 
    title = "Avg House Price (Lakhs)",
    position = "bottomright"
  )

```

```{r}
# Calculate average price per city
city_summary <- dt %>%
  group_by(City) %>%
  summarise(
    Median_Price_Lakhs = median(Price_in_Lakhs, na.rm = TRUE),
    Avg_Price_Lakhs = mean(Price_in_Lakhs, na.rm = TRUE),
    Property_Count = n(),
    # Calculate percentages for each direction
    Pct_North = round(mean(Facing == "North", na.rm = TRUE) * 100, 1),
    Pct_South = round(mean(Facing == "South", na.rm = TRUE) * 100, 1),
    Pct_East  = round(mean(Facing == "East", na.rm = TRUE) * 100, 1),
    Pct_West  = round(mean(Facing == "West", na.rm = TRUE) * 100, 1)
  )

city_geo$Median_Price_Lakhs = city_summary$Median_Price_Lakhs
city_geo <- city_geo %>%
  select(City, latitude, longitude,) %>% # Keep only geo columns from the old city_geo
  left_join(city_summary, by = "City")

#Creating interactive Map
library(leaflet)

pal <- colorNumeric(
  palette = "RdYlGn",
  domain = city_geo$Median_Price_Lakhs,
  reverse = TRUE
)

leaflet(city_geo) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude, 
    lat = ~latitude,
    radius = 5,
    weight = 1,
    fillOpacity = 0.8,
    color = "white",
    fillColor = ~pal(Median_Price_Lakhs),
    popup = ~paste0("<b>City:</b> ", City, 
                    "<br><b>Median Price:</b> ", round(Median_Price_Lakhs, 2), 
                    " Lakhs","<br><b>Total Properties:</b> ", Property_Count,
      "<hr style='margin: 5px 0;'>",
      "<b>Orientation Distribution:</b><br>",
      "North: ", Pct_North, "% | South: ", Pct_South, "%<br>",
      "East: ", Pct_East, "% | West: ", Pct_West, "%"),
    clusterOptions = markerClusterOptions()
  ) %>%
  # Add a legend to make the colors clear
  addLegend(
    pal = pal, 
    values = ~Median_Price_Lakhs, 
    opacity = 0.7, 
    title = "Median House Price (Lakhs)",
    position = "bottomright"
  )
```

```{r}
#Zone Wise Data

library(dplyr)

dt_zoned <- dt %>%
  mutate(Zone = case_when(
    City %in% c("New Delhi", "Ludhiana", "Jaipur", "Jodhpur", "Lucknow", 
                "Dehradun", "Noida", "Faridabad", "Amritsar", "Haridwar", 
                "Gurgaon","Dwarka") ~ "North",
    
    City %in% c("Chennai", "Coimbatore", "Bangalore", "Warangal", "Hyderabad", 
                "Vishakhapatnam", "Mysore", "Vijayawada", "Trivandrum", 
                "Kochi", "Mangalore") ~ "South",
    
    City %in% c("Mumbai", "Pune", "Nagpur", "Ahmedabad", "Surat") ~ "West",
    
    City %in% c("Kolkata", "Ranchi", "Durgapur", "Gaya", "Jamshedpur", 
                "Patna", "Cuttack", "Bhubaneswar") ~ "East",
    
    City %in% c("Bhopal", "Indore", "Raipur", "Bilaspur") ~ "Central",
    
    City %in% c("Guwahati", "Silchar") ~ "Northeast",
    
    TRUE ~ "Other"
  ))

zone_summary <- dt_zoned %>%
  group_by(Zone) %>%
  summarise(
    Median_Price_Lakhs = median(Price_in_Lakhs, na.rm = TRUE),
    House_Count = n()
  ) %>%
  arrange(desc(Median_Price_Lakhs))


bp = barplot(zone_summary$Median_Price_Lakhs, 
        names.arg = zone_summary$Zone, 
        col = "darkorange",
        main = "Median House Price by Zones",
        ylab = "Price (in Lakhs)",
        las = 1, xpd = FALSE,
        ylim = c(245, 260))



text(x = bp, 
     y = zone_summary$Median_Price_Lakhs, 
     labels = round(zone_summary$Median_Price_Lakhs, 2), 
     pos = 3,         
     cex = 0.9, xpd = FALSE,      
     font = 2,        
     col = "black")
```

### States with the Most Affordable Housing

```{r}
state_summary <- dt %>%
  group_by(State) %>%
  summarise(
    Median_Price_Lakhs = median(Price_in_Lakhs, na.rm = TRUE),
    Avg_Price_Lakhs = mean(Price_in_Lakhs, na.rm = TRUE),
    Property_Count = n()
  )%>% arrange(Median_Price_Lakhs)
state_summary
```

-   States appearing at the top of the list offer comparatively more affordable housing options, while states toward the bottom exhibit higher average prices, often associated with stronger urban demand.

### Schools

```{r}
cor(dt_expanded$Price_in_Lakhs, dt_expanded$Nearby_Schools)
cor(dt_expanded$Price_in_Lakhs, dt_expanded$Nearby_Hospitals)
```

### Floors

```{r}
cor(dt$Floor_No,dt$Total_Floors)
```

-   We can observe that floor no. and total no. of floors have significantly low correlation.

### Locality wise grouping

```{r}
#| echo: true
locality_luxury <- dt %>%
  group_by(Locality, City) %>%
  summarise(
    Avg_Price = mean(Price_in_Lakhs, na.rm = TRUE),
    Median_Price = median(Price_in_Lakhs),
    Sample_Size = n()
  ) %>% 
  arrange(desc(Median_Price)) 

print(locality_luxury)


locality_to_join <- locality_luxury %>%
  select(Locality, City, Median_Price_Locality = Median_Price)

dt_expanded <- dt_expanded %>%
  left_join(locality_to_join, by = c("Locality", "City"))

#Confirming that there are no missing values.
sum(is.na(dt_expanded$Median_Price_Locality))
```

## Hypothesis Testing

### Research Statement: The prices of villas are generally higher than the apartments.

-   **NULL HYPOTHESIS**:- The average prices of the apartments are equal to the average prices of the villas.

-   **ALTERNATIVE HYPOTHESIS**:- The average prices of the villas are higher than the average prices of the apartments.

-   **H0:** mean_price_of_villa = mean_price_of_apartment

-   **H1:** mean \_price_of_villa \> mean_price_of_apartment

    ```{r}
    t.test(Price_in_Lakhs ~ Property_Type,
           data = subset(dt, Property_Type %in% c("Villa","Apartment")),
           alternative = "greater")

    ```

**Interpretation**:-

Here the p-values are greater than 0.05, which implies that we failed to reject the null hypothesis. Thus, based on our data-set, we cannot conclude that the prices of villas are higher than that of apartments.

There is insufficient evidence to conclude that the average price of villas are higher than that of apartments.

### Research Statement :- Do Furnished homes have more cost than the Unfurnished homes ?

-   **NULL HYPOTHESIS**:- The furnished and unfurnished houses have the same price.

-   **ALTERNATIVE HYPOTHESIS**:- The furnished houses have higher prices than the unfurnished houses

    ```{r}
    subset_dt <- subset(dt, Furnished_Status %in% c("Furnished", "Unfurnished"))

    t.test(Price_in_Lakhs ~ Furnished_Status,
           data = subset_dt)
    ```

**Interpretation:**

Since the p-values = 0.44 \>0.05, we fail to reject the null hypothesis. Thus, we can say that there is no significant difference between the prices of the houses which are furnished or unfurnished(based on the data).

## Feature Engineering

```{r}
names(Filter(is.character, dt_expanded))
```

```{r}

#Columns: Property_Type,Furnished_Status,Facing, Owner_Type, Availability_Status.
dt_expanded <- dummy_cols(dt_expanded, 
                          select_columns =c( "Property_Type","Furnished_Status"
                                             ,"Facing","Owner_Type",
                                             "Availability_Status"), 
                          remove_selected_columns = TRUE,
                          remove_first_dummy = TRUE)


#Column: Public_Transport_Accessibility.
dt_expanded <- dt_expanded %>%
  mutate(Public_Transport_Accessibility = case_when(
    Public_Transport_Accessibility == "Low" ~ 0,
    Public_Transport_Accessibility == "Medium" ~ 1,
    Public_Transport_Accessibility == "High" ~ 2
  ))
```

-   Now our Data is ready for Modelling.

    ```{r}
    #Preventing data loss
    dtm = dt_expanded
    ```

## Model Training

```{r}
#Checking for missing values
sum(is.na(dtm))
```

-   Our data has no missing value.

```{r}

library(dplyr)
library(car)

# Create the modeling dataset by dropping non-numeric and redundant columns
X_model <- dtm %>%
  dplyr::select(
    -ID, -State, -City, -Locality,   # Character columns
    -Year_Built                       
  )
```

### Multicollinearity Check

```{r}
#| warning: false
#  Build a preliminary model
prelim_model <- lm(Price_in_Lakhs ~ ., data = X_model)

#  Calculate VIF 
vif_results <- data.frame(VIF = vif(prelim_model)) %>%
  tibble::rownames_to_column("Feature")%>%
  arrange(desc(VIF))

print(head(vif_results,5))
```

-   VIF values of all columns are \<5; which signifies that the columns are not linearly dependent.

### Multiple Linear regression

```{r}
library(caret)


set.seed(36)
trainIndex <- createDataPartition(X_model$Price_in_Lakhs, p = 0.75, list = FALSE)
train_data <- X_model[trainIndex, ]
test_data  <- X_model[-trainIndex, ]


lm1 <- lm(Price_in_Lakhs ~ ., data = train_data)


summary(lm1)

```

```{r}
lm2 <- lm(Price_in_Lakhs ~ Median_Price_Locality+Size_in_SqFt+Price_per_SqFt+Security+Property_Type_Villa , data = train_data)

summary(lm2)
```

-   We tried multiple Test-train ratios (ratio = 0.7, 0.75, 0.8, 0.9). Among them the **Average Multiple R-squared value** is 0.5103. It means **around 51% variations in the data can be explained on the basis of this Multiple Linear Regression Model.**

```{r}
#Checking Normal Q-Q Plot to find model's errors follow a normal distribution or not.
qqnorm(lm2$residuals)
qqline(lm2$residuals, col="red")
```

-   The points follow the red diagonal line in the middle but **"bow" significantly at both ends** (the tails).
-   **Verdict**: This "S-shape" suggests that our data has a skewed distribution which a standard linear model isn't fully accounting for.

### Generalized Linear Model

```{r}
set.seed(36)
trainIndex <- createDataPartition(X_model$Price_in_Lakhs, p = 0.75, list = FALSE)
train_data <- X_model[trainIndex, ]
test_data  <- X_model[-trainIndex, ]

lm2 <- lm(Price_in_Lakhs ~ Median_Price_Locality+Size_in_SqFt+Price_per_SqFt+Security+Property_Type_Villa , data = train_data)

# GLM with Gamma distribution and Log Link
glm1 <- glm(Price_in_Lakhs ~ Median_Price_Locality+Size_in_SqFt+Price_per_SqFt+Property_Type_Villa, 
                 family = Gamma(link = "log"), 
                 data = train_data)
summary(glm1)


# Comparing the AIC (Lower is better)
AIC(lm2)
AIC(glm1)
```

```{r}
par(mfrow = c(1,2))
qqnorm(lm2$residuals,main = "Normal Q-Q Plot for LM model")
qqline(lm2$residuals, col="red")
qqnorm(glm1$residuals,main = "Normal Q-Q Plot for GLM model")
qqline(glm1$residuals, col="magenta")
```

-   We can observe that the **GLM** seems to be more robust and statistically appropriate model for the Indian housing market.

-   We can see that in the middle section (between -2 and 2 on the x-axis), the points hug the diagonal line much more tightly than they did in the LM plot.

-   While there is still some "S-shaping" at the very extreme ends, the overall "bend" is more symmetrical. This shows that the **Gamma distribution** is doing a much better job of modeling the skewed nature of real estate prices.

```{r}
# Predict using both models on our test data
pred_lm  <- predict(lm2, newdata = test_data)
pred_glm <- predict(glm1, newdata = test_data, type = "response")

# Calculate Root Mean Squared Error (RMSE)
rmse_lm  <- sqrt(mean((test_data$Price_in_Lakhs - pred_lm)^2))
rmse_glm <- sqrt(mean((test_data$Price_in_Lakhs - pred_glm)^2))

# RMSE
cat("LM RMSE:", rmse_lm, "Lakhs\n")
cat("GLM RMSE:", rmse_glm, "Lakhs")
```

-   While the **GLM** provided a better statistical fit for the non-linear relationship between size and price, it proved to be highly sensitive to outliers during prediction. The **LM**, despite violating some normality assumptions, achieved a much lower **RMSE (99.04 Lakhs)** compared to the **GLM (422.25 Lakhs)**, making it the more reliable model for actual price estimation in this dataset.

```{r}
# Calculate Mean Absolute Error (MAE)
mae_lm  <- mean(abs(test_data$Price_in_Lakhs - pred_lm))
mae_glm <- mean(abs(test_data$Price_in_Lakhs - pred_glm))

cat("LM MAE:", mae_lm, "\n")
cat("GLM MAE:", mae_glm)
```

-   Our evaluation shows that while the **Linear Model (LM)** remains the most accurate with an MAE of **79.15 Lakhs**, the **GLM (Gamma)** provides a more theoretically sound fit for the data distribution. The disparity between the **GLM's RMSE (422.25)** and **MAE (128.95)** indicates that the model is performing well on the majority of the dataset but is highly sensitive to a small number of cases.

### Decision Tree

```{r}
library(rpart)
library(rpart.plot)

tree_model <- rpart(Price_in_Lakhs ~ Median_Price_Locality + Size_in_SqFt + 
                    Security + Property_Type_Villa, 
                    data = train_data, 
                    method = "anova",
                    control = rpart.control(cp = 0.00001, maxdepth = 4))


prp(tree_model, extra = 1, box.palette = "RdYlGn", main = "Multi-Variable House Price Tree")

```

-   The model has determined that **Median_Price_Locality** is so dominant that it has used it for every single split in the tree, completely ignoring Size, Security, and Property Type for now.

```{r}
#Extract importance scores
importance_scores <- tree_model$variable.importance


library(dplyr)

import_df <- data.frame(Importance = importance_scores) %>%
  tibble::rownames_to_column("Feature") %>%
  mutate(Percentage = (Importance / sum(Importance)) * 100) %>%
  arrange(desc(Percentage))
import_df
```

-   Our Variable Importance analysis reveals that 'Location' is the primary determinant of property value in our dataset, accounting for over 99.9% of the model's predictive power. While physical attributes like 'Size' are traditionally important, the sheer variance in price between different Indian localities makes the specific neighborhood the only statistically significant predictor for our current decision tree.

```{r}
# Predict on Test Data
pred_tree <- predict(tree_model, newdata = test_data)

# Calculate Errors
rmse_tree <- sqrt(mean((test_data$Price_in_Lakhs - pred_tree)^2))
mae_tree  <- mean(abs(test_data$Price_in_Lakhs - pred_tree))

cat("Decision Tree MAE:", mae_tree, "Lakhs\n")
cat("Decision Tree RMSE:", rmse_tree, "Lakhs")
```

-   The **Decision Tree** highlights that location is the absolute driver of value, with **Median_Price_Locality** holding a **99.94% importance score**. While its higher **MAE of 115.27 Lakhs** shows it is currently less precise for individual predictions than our Linear Model.

### Random Forest

-   Since our **RMSE** and **MAE** results revealed that luxury outliers and non-linear patterns are challenging for our **LM** and **GLM**, a **Random Forest** is the natural evolution for our project. It builds hundreds of decision trees and averages them, which should significantly reduce the high errors.

```{r}
#| warning: false
library(randomForest)

set.seed(12)
train_sample <- train_data[sample(1:nrow(train_data), 30000), ]


rf_model <- randomForest(Price_in_Lakhs ~ Median_Price_Locality + Size_in_SqFt + 
                         Security + Property_Type_Villa, 
                         data = train_sample, 
                         ntree = 400, 
                         importance = TRUE)


pred_rf <- predict(rf_model, newdata = test_data)
rmse_rf <- sqrt(mean((test_data$Price_in_Lakhs - pred_rf)^2))
mae_rf  <- mean(abs(test_data$Price_in_Lakhs - pred_rf))

cat("Random Forest MAE:", mae_rf, "Lakhs\n")
cat("Random Forest RMSE:", rmse_rf, "Lakhs")
```

-   <div>

    <div>

    The **Linear Model (LM)** is the most effective tool for this dataset, achieving the lowest average error with a **Mean Absolute Error (MAE) of 79.15 Lakhs**. While more complex models like the **GLM** and **Random Forest** were explored, the LM's ability to provide stable predictions across the price spectrum makes it the most reliable "calculator" for our final results.

    </div>

    </div>
